The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Slight wording changes

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most MWA observations are recorded at a resolution of 0.5~s and 40~kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$ values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations for each observing night. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global RFI fraction in the 72.3--230.8 MHz range.

>>>>

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to apply the RFI detection from the first run and convert the raw files to his/her preferred output.

When time or frequency averaging is requested, \textsc{cotter} averages samples together that have not been flagged. When all input samples are flagged, the average of all input visibilities is stored into the output sample, and the output sample is flagged. This method makes it possible to superficially analyse flagged samples in the output, even though information is lost in the averaging. \textsc{cotter} stores a weight for each visibility in the output file, which is scaled to the number of unflagged input samples that were used for the output sample. Because of this, when no averaging is requested, the output is 50\% larger than the input (one extra float per complex float visibility). In practice, most MWA observations are recorded at a resolution of 0.5~s and 40~kHz, and averaged to a resolution of 2--4 s and 40--80 kHz to reduce the data volume. This does not cause any significant time or bandwidth decorrelation up to the first beam null for MWA data. \textsc{cotter} performs the phase shifting and cable delay corrections before averaging, and recalculates the $u,v,w$-values for the central time and frequency of the output sample. This helps to prevent time/frequency decorrelation.

@@

\section{Detection results} \label{ch:detection-results}
The ``RFI'' column of Table~\ref{tbl:obs-list} lists the fraction of samples that were classified as RFI by the \textsc{aoflagger} in the cross-correlations for each observing night. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% of detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in a global RFI occupancy of 1.13\% in the 72.3--230.8 MHz range.
